Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-09-28 pnas/Towards.tex
small modifications

\begin{figure}[h]
	\centering
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[angle=270, width=.7\textwidth]{TRPV1_rec_incl_noisy.eps}
		\caption{TPRV1 with cutoff $L_{\text{max}}=5$}\label{fig:TRPV1_recon}
	\end{subfigure}
	\hfill
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[angle=270, width=.7\textwidth]{BPTI_maxL2_L20.eps}
		\caption{BPTI with cutoff $L_{\text{max}}=2$}\label{fig:BPTI_recon}
	\end{subfigure} 
	\caption{\label{fig:cryo_recon} Reconstructions from the first, second and third order autocorrelations.
		The ground truth volumes were expanded according to~\eqref{eq:volume_expansion_app} with cutoff $L_{\text{max}}$.
		\revised{The original molecules are shown in purple and the (smoothed) ground truths in blue  to illustrate the smoothing effect of our downsampling and truncation of the spherical harmonics expansion.
		The reconstructions  from clean autocorrelations are shown in yellow and recovery from 
		autocorrelations estimated from noisy data in gray. For the noisy experiments, we used 300 micrographs with SNRs of 1/16 for TRPV1 and 1/64 for BPTI. We present excerpts from noisy micrographs side-by-side with the corresponding clean ones in Figure~\ref{fig:cryo_noisy_micros_for_rec}}.}
		 \vspace{-20pt} 	
\end{figure}
% -------------- Vertical subfigures: -------------------------------

@@

\begin{figure*}[t!]
	\centering
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{TRPV1_clean_micro.png}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{TRPV1_noisy_micro_SNR_16.png}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{clean_micro_BPTI.png}
		\caption{}
	\end{subfigure} \hfill
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{noisy_micro_BPTI_SNR_64.png}
		\caption{}
	\end{subfigure}
	\caption{\label{fig:cryo_noisy_micros_for_rec} \revised{Excerpts from the noisy micrographs used for the reconstructions in Figure~\ref{fig:cryo_recon} and the corresponding clean excerpts. (a) Excerpt of the clean TRPV1 micrograph; (b) Excerpt of the noisy TRPV1 micrograph, with SNR $= 1/16$; (c) Excerpt from the clean BPTI micrograph; (d) Excerpt from the noisy BPTI micrograph with SNR $= 1/64$.}} 
\end{figure*}

\begin{figure*}[t!]
	\centering
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\includegraphics[scale=0.35]{TRPV1_trunc_FSC.eps}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\includegraphics[scale=0.35]{BPTI_FSC.eps}
		\caption{}
	\end{subfigure} 
	\caption{\label{fig:cryo_FSC} \revised{Fourier Shell Correlations (FSCs) for the noisy reconstructions presented in Figure~\ref{fig:cryo_recon} using the $0.5$ criterion, compared to the (low-resolution) ground truth. (a) FSC for the TRPV1 reconstruction, giving resolution of $24 \AA$; (b) FSC for the BPTI reconstruction, giving resolution of $13 \AA$.}} 
\end{figure*}

@@

For the \revised{clean} experiments in Figure~\ref{fig:cryo_recon}, we computed the analytical first three autocorrelations of the volume as explained in Appendix~\ref{sec:ac_cryo}.
To estimate the coefficients of the volume itself, we solve the LS problem
\begin{equation} \label{eq:LS_cryo}
\min_{\hat{x},\hat{\gamma}} w_1 \vert \mathfrak{a}_x^1 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^1 \vert^2 + w_2\| \mathfrak{a}_x^2 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^2 \|_2^2 + w_3\| \mathfrak{a}_x^3 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^3 \|_F^2, 
\end{equation}  
where the explicit expressions of $\mathfrak{a}_x^1$, $\mathfrak{a}_x^2$ and $\mathfrak{a}_x^3$ are given in~\eqref{eq:ac_1_prolates},~\eqref{eq:ac_2_prolates} and~\eqref{eq:ac_3_prolates}, respectively.  
In the experiments, we set $w_1=w_2=w_3=1$. 
The LS problem was solved using Matlab's \texttt{lsqnonlin} solver for nonlinear LS problems. The expected $\hat \gamma$ in this experiment is one.

@@

The true volume used in the experiments in Figures~\ref{fig:cryo_detection} and~\ref{fig:BPTI_recon} was the Bovine Pancreatic Trypsin Inhibitor (BPTI) mutant with altered binding loop sequence, whos atomic model is available in the Protein Data Bank (PDB) %~\cite{pdb} 
as 1QLQ\footnote{\url{https://www.rcsb.org}}. We generated an EM map from this atomic model in UCSF Chimera~\cite{pettersen2004ucsf} at a resolution of 5~\AA, and cropped it to remove zeros at the boundary to obtain a volume of size $31^3$. For the experiment in Figure~\ref{fig:BPTI_recon}, the volume was downsampled to size $20^3$. For the experiment in Figure~\ref{fig:TRPV1_recon}, we used the TRPV1 in complex with DkTx and RTX, whose EM map is available in the  Electron Microscopy Data Bank (EMDB) %~\cite{emdb} 
as EMD-8117\footnote{\url{http://www.ebi.ac.uk/pdbe/emdb}}. The original map has size $192^3$, and was downsampled to size $20^3$. To generate the ground truth for our reconstructions, we expanded both volumes as in~\eqref{eq:volume_expansion_app} with cutoff $L_{\text{max}} = 5$ for TRPV1 and $L_{\text{max}} = 2$ for BPTI. For the TRPV1 reconstruction, the optimizer converged to a point giving relative $\ell_2$ error of $10^{-6}$ in the first three autocorrelations and an error of $10^{-1}$ in the expansion coefficients of the volume. For the BPTI reconstruction, the errors in the autocorrelations were $10^{-6}$, $10^{-7}$ and $10^{-8}$ for the third, second and first autocorrelations, respectively, while the error in the expansion coefficients of the volume was $5\times 10^{-2}$. This illustrates the ill-conditioning of the map between the volume and its first three autocorrelations that prevents us from obtaining results from noisy data.

>>>>

\begin{figure}[h]
	\centering
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[angle=270, width=.7\textwidth]{TRPV1_rec_incl_noisy.eps}
		\caption{TPRV1 with cutoff $L_{\text{max}}=5$}\label{fig:TRPV1_recon}
	\end{subfigure}
	\hfill
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[angle=270, width=.7\textwidth]{BPTI_maxL2_L20.eps}
		\caption{BPTI with cutoff $L_{\text{max}}=2$}\label{fig:BPTI_recon}
	\end{subfigure} 
	\caption{\label{fig:cryo_recon} Reconstructions from the first, second and third order autocorrelations.
		The ground truth volumes were expanded according to~\eqref{eq:volume_expansion_app} with cutoff $L_{\text{max}}$.
		\revised{The original molecules are shown in purple and the (smoothed) ground truths in blue  to illustrate the smoothing effect of our downsampling and truncation of the spherical harmonics expansion.
		The reconstructions  from clean autocorrelations are shown in yellow and recoveries from 
		autocorrelations estimated from noisy data in gray. For the noisy experiments, we used 300 micrographs with SNRs of 1/16 for TRPV1 and 1/64 for BPTI. We present excerpts from noisy micrographs side-by-side with the corresponding clean ones in Figure~\ref{fig:cryo_noisy_micros_for_rec}}.}
		 \vspace{-20pt} 	
\end{figure}
% -------------- Vertical subfigures: -------------------------------

@@

\begin{figure*}[t!]
	\centering
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{TRPV1_clean_micro.png}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{TRPV1_noisy_micro_SNR_16.png}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{clean_micro_BPTI.png}
		\caption{}
	\end{subfigure} \hfill
	\begin{subfigure}[t]{0.245\textwidth}
		\centering
		\includegraphics[scale=0.3]{noisy_micro_BPTI_SNR_64.png}
		\caption{}
	\end{subfigure}
	\caption{\label{fig:cryo_noisy_micros_for_rec} \revised{Excerpts from the noisy micrographs used for the reconstructions in Figure~\ref{fig:cryo_recon} and the corresponding clean excerpts. (a) Excerpt of the clean TRPV1 micrograph; (b) Excerpt of the noisy TRPV1 micrograph with SNR $= 1/16$; (c) Excerpt from the clean BPTI micrograph; (d) Excerpt from the noisy BPTI micrograph with SNR $= 1/64$.}} 
\end{figure*}

\begin{figure*}[t!]
	\centering
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\includegraphics[scale=0.35]{TRPV1_trunc_FSC.eps}
		\caption{}
	\end{subfigure} \hfill %\hspace{2pt}
	\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\includegraphics[scale=0.35]{BPTI_FSC.eps}
		\caption{}
	\end{subfigure} 
	\caption{\label{fig:cryo_FSC} \revised{Fourier Shell Correlations (FSCs) for the noisy reconstructions presented in Figure~\ref{fig:cryo_recon} using the $0.5$ criterion, compared to the (smoothed) ground truth. (a) FSC for the TRPV1 reconstruction, giving resolution of $24 \AA$; (b) FSC for the BPTI reconstruction, giving resolution of $13 \AA$.}} 
\end{figure*}

@@

For the \revised{clean} experiments in Figure~\ref{fig:cryo_recon}, we computed the analytical first three autocorrelations of the volume as explained in Appendix~\ref{sec:ac_cryo}.
To estimate the coefficients of the volume itself, we solve the LS problem
\begin{equation} \label{eq:LS_cryo}
\min_{\hat{x},\hat{\gamma}} w_1 \vert \mathfrak{a}_x^1 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^1 \vert^2 + w_2\| \mathfrak{a}_x^2 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^2 \|_2^2 + w_3\| \mathfrak{a}_x^3 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^3 \|_\text{F}^2, 
\end{equation}  
where the explicit expressions of $\mathfrak{a}_x^1$, $\mathfrak{a}_x^2$ and $\mathfrak{a}_x^3$ are given in~\eqref{eq:ac_1_prolates},~\eqref{eq:ac_2_prolates} and~\eqref{eq:ac_3_prolates}, respectively.  
In the experiments, we set $w_1=w_2=w_3=1$. 
The LS problem was solved using Matlab's \texttt{lsqnonlin} solver for nonlinear LS problems. The expected $\hat \gamma$ in this experiment is one.

@@

The true volume used in the experiments in Figures~\ref{fig:cryo_detection} and~\ref{fig:BPTI_recon} was the Bovine Pancreatic Trypsin Inhibitor (BPTI) mutant with altered binding loop sequence, whos atomic model is available in the Protein Data Bank (PDB) %~\cite{pdb} 
as 1QLQ\footnote{\url{https://www.rcsb.org}}. We generated an EM map from this atomic model in UCSF Chimera~\cite{pettersen2004ucsf} at a resolution of 5~\AA, and cropped it to remove zeros at the boundary to obtain a volume of size $31^3$. For the experiment in Figure~\ref{fig:BPTI_recon}, the volume was downsampled to size $20^3$. For the experiment in Figure~\ref{fig:TRPV1_recon}, we used the TRPV1 in complex with DkTx and RTX, whose EM map is available in the  Electron Microscopy Data Bank (EMDB) %~\cite{emdb} 
as EMD-8117\footnote{\url{http://www.ebi.ac.uk/pdbe/emdb}}. The original map has size $192^3$, and was downsampled to size $20^3$. To generate the ground truth for our reconstructions, we expanded both volumes as in~\eqref{eq:volume_expansion_app} with cutoff $L_{\text{max}} = 5$ for TRPV1 and $L_{\text{max}} = 2$ for BPTI. For the TRPV1 reconstruction, the optimizer converged to a point giving relative $\ell_2$ error of $10^{-6}$ in the first three autocorrelations and an error of $10^{-1}$ in the expansion coefficients of the volume. For the BPTI reconstruction, the errors in the autocorrelations were $10^{-6}$, $10^{-7}$ and $10^{-8}$ for the third, second and first autocorrelations, respectively, while the error in the expansion coefficients of the volume was $5\times 10^{-2}$. This illustrates the ill-conditioning of the map between the volume and its first three autocorrelations that prevents us from obtaining \revised{high-resolution} results from noisy data.
